Statistical Inference for inter-arrival times of extreme events in bursty time series

### 2020-08-02 article_springer/article_springer.Rmd
$$ --> align

  A new model for the _inter-exceedance times_ of events 
  above high thresholds is proposed; the threshold exceedances itself are modeled via the standard 
  Peaks-Over-Threshold (POT) method. For high thresholds and 
  infinite-mean waiting times, it is shown that the times between threshold 
  crossings are Mittag-Leffler distributed, and thus form a "fractional 
  Poisson Process" which generalizes the standard Poisson Process of threshold
  exceedances.
  Graphical means of estimating model parameters and assessing 
  model fit are provided. Along the way, the inference method gets applied to a real-world
  bursty time series, and it is shown how the memory of the Mittag-Leffler distribution affects 
  the predictive distribution for the time until the next extreme event.

@@

We review the marked renewal process in Section 2, and 
derive a scaling limit theorem for inter-exceedance times in Section 3.
We give a statistical procedure to estimate model parameters via stability plots 
in Section 5, but to set the stage we first discuss inference for the 
Mittag-Leffler distribution in Section 4. A simulation study of the effectiveness of our statistical procedure is given in Section 6. In Section 7 we apply our method to a real data set.  In Section 8, we discuss the memory property of the  Mittag-Leffler distribution, and how it affects the predictive distribution for the time until the next threshold crossing event. Finally we close with a discussion and conclusion in Section 9.
For all statistical computations we have used R [@R]. 
All code and data used for the analysis in this article has been organized into 
an R package `CTRE` (<https://github.com/strakaps/CTRE>).
The source code for the figures generated in this manuscript is 
available online at <https://github.com/strakaps/bursty-POT>.

# Continuous Time Random Exceedances (CTRE)

@@

: Let $(W,J), (W_1, J_1), (W_2, J_2), \ldots$ be i.i.d. pairs of random variables, where the $W_k > 0$ are interpreted as the *waiting times* 
and $J_k \in \mathbb{R}$ as the *event magnitudes*. If $W$ and $J$ are independent, the Marked Renewal Process is said to be *uncoupled*.

@@

: Given a threshold $\ell \in (x_L, x_R)$, 
consider the stopping time 
$$\tau(\ell) := \min\{k: J_k > \ell\},\quad \ell \in (x_L, x_R).$$
Define the pair of random variables $(X(\ell), T(\ell))$ via 
$$X(\ell) = J_{\tau(\ell)} - \ell, \quad 
T(\ell) = \sum_{k=1}^{\tau(\ell)} W_k.$$
By restarting the MRP at $\tau(\ell)$, inductively define the
two i.i.d. sequences $T(\ell,n)$ and $X(\ell, n)$, $n \in \mathbb N$, 
called the "inter-arrival times" and the "exceedances", respectively.
The pair sequence $(X(\ell, n), T(\ell, n))_{n \in \mathbb N}$ is called 
a Continuous Time Random Exceedance model (CTRE). 
If the underlying MRP is uncoupled, then the CTRE is also called 
uncoupled.

@@

# Scaling limit of Exceedance Times {#sec:scaling}

@@

$$
\frac{f(\lambda x)}{f(x)} = \lambda^\rho \quad \text{ for all } \lambda > 0.
$$

For more details regular variation and stable limit theorems, we recommend the book by @MeerschaertSikorskii.

@@

: For the magnitudes $J_k$ assumption \eqref{assumptionJs} holds. Furthermore, let the waiting times $W_k$ be in the domain of attraction of a positively skewed sum-stable law with stability parameter $0 < \beta < 1$; more precisely,
\begin{align} \label{eq:stability}
\frac{W_1 + \ldots + W_n}{b(n)} \overset{d}{\longrightarrow} D, 
\quad n \to \infty
\end{align}
for a function $b(n)$ which is regularly varying at $\infty$ with parameter $1/\beta$, and where $\mathbf E[\exp(-sD)] = \exp(-s^\beta)$, $s > 0$. Write $p_{\ell} := \mathbf P(J > \ell)$. Then the weak convergence
$$
\frac{T(\ell)} {b(1/p_{\ell})} \to Z_\beta \quad \text{ as } \quad \ell \uparrow x_R
$$
holds, where the Mittag-Leffler random variable $Z_\beta$ is defined on 
the positive real numbers via 
$$
\mathbf E[\exp(-sZ_\beta)] = \frac{1}{1+s^\beta}.
$$

*Proof of Theorem:* 	Due to assumption \eqref{assumptionJs}, it follows that
	\begin{align*}
	P\left(\frac{M(c)-d(c)}{a(c)} \leq x\right) \rightarrow F(x) := F(x,\xi) \text{ as } c \rightarrow \infty,
	\end{align*}
	where $x$ is any value from the support of $F$. Then
	\begin{align*}
	  F(xa(c)+d(c))^{\lfloor c \rfloor} &\rightarrow F(x) \text{ as } c \rightarrow \infty,\\
      \Rightarrow  \lfloor c \rfloor \log F(\ell_c)  &\rightarrow \log F(x)\text{ as } c \rightarrow \infty,
	\end{align*}
	with $\ell_c:=xa(c)+d(c)$. Furthermore, since $\log(1-x) \sim -x$ for small $x$ it follows
    \begin{align}
      c \cdot p(\ell_c) \rightarrow -\log F_A(x)  \text{ as } c \rightarrow \infty,\label{eq:lplconv}
    \end{align}
	 with $p(\ell_c):= 1-F(\ell_c)$. Since $\tau(\ell_c) \sim Geo(p(\ell_c))$ and \eqref{eq:lplconv}
	it follows that $\tau(\ell_c)/c$ converges to an Exponential distributed random variable,
	\begin{align*}
	\frac{\tau(\ell_c)}{c} \Rightarrow E  \text{ as } c \rightarrow \infty
	\end{align*}
	 with inverse mean $\lambda:=-\log F_A(x)$. Due to
	\begin{align*}
	S(c):= \sum_{i=1}^{\lfloor c \rfloor} \frac{W_i}{b(c)} \Rightarrow D \ \text{ as } c \rightarrow \infty,
	\end{align*}
    it follows with Gnedenko's transfer theorem (see @gnedenko1983limit), that
	\begin{align}
	\sum_{i=1}^{ \tau(\ell_c)} \frac{W_i}{b(\ell_c)}  \Rightarrow Z\ \text{ as }  c \rightarrow \infty, \label{TheoGutArg}
	\end{align}
	where the distribution of $Z$ has the characteristic function
	\begin{align}
    \Psi(s) &= \int_0^{\infty} (\Phi_D(s))^{y} F_E(dy) \nonumber\\
    &= \int_0^{\infty} \frac{1}{1-\log(\Phi_D(s))/\lambda} \nonumber\\
    &= \frac{1}{1+s^{\beta}\lambda^{-1}}
	\end{align}
    where $\Phi_D(s) = \exp(-s^\beta)$ is the Laplace transform of $D$ 
    and  $F_E(dy)$ is the distribution function of $E$. Hence, $Z$ is Mittag-Leffler distributed with parameter $\lambda^{-1/\beta}$. Rewriting
	\begin{align*}
    \sum_{i=1}^{ \tau(\ell_c)} \frac{W_i}{b(1/p(\ell_c))} = \left(\sum_{i=1}^{ \tau(\ell_c)} \frac{W_i}{b(c) }\right) \frac{b(c)}{b(1/p(\ell_c))}  
	\end{align*}
	and
	\begin{align*}
	\frac{b(c)}{b(1/p(\ell_c))} = \frac{b(c)}{b(c/(p(\ell_c)c))} \sim (p(\ell_c) \cdot c)^{1/\beta} \rightarrow \lambda^{1/\beta} \text{ as } c \rightarrow \infty,
	\end{align*}
	the second factor converges to $\lambda^{1/\beta}$, and it follows that
	 $Z \sim ML(\beta,1)$. Since $c \rightarrow \infty$ is equivalent to $\ell \uparrow x_R$, the assertion follows with $\ell:=\ell_c$ and $p_\ell:=p(\ell_c)$.
\qed

@@

: If $\beta = 1$, the result of the Theorem above is standard, see e.g. 
Equation (2.2) in @Gut1999. In @Anderson1987  a similar result is shown with a different choice of scaling constant.

@@

# Model Choice and inference for the Mittag-Leffler distribution{#sec:ML}

The classical POT approach is based on the fact that exceedances above a high threshold are asymptotically GPD distributed, hence a GPD distribution is fitted to the exceedances above a high enough threshold. As this approach is by now well-established in extreme value theory, we do not discuss it any further. Instead, this article focuses on our new contribution: the modelling of inter-exceedance times. For literature pointing to POT see e.g. @smith1984threshold; @leadbetter1991basis; @davison1990models; @beirlantBook Ch.5.3; @embrechts2013modelling Ch. 6.5.; @ColesBook Ch.4.

Due to the Theorem in the last Section we know that for high thresholds the times between exceedances are asymptotically Mittag-Leffler distributed in case of heavy-tailed waiting times between the observations. In this article we propose to model inter-exceedance times in a similar fashion as magnitudes in the classical POT approach, more precisely to fit a Mittag-Leffler distribution to the inter-exceedance times via stability plots. Hence, before we discuss inference for the inter-exceedance times in the next section, we first discuss inference for Mittag-Leffler distributions.

Due to the form of the density function of a Mittag-Leffler distribution there exists no closed form for the Maximum Likelihood estimator (MLE). In the R Package `MittagLeffleR` [@MittagLeffleR], Maximum Likelihood estimation is implemented via numerical optimization. The MLE slightly outperforms the log-Moment estimator regarding bias and RMSE for big enough sample sizes, but is extremely computationally intensive. Figure \ref{fig:MSE} shows that both estimators perform well, even for small sample sizes. For smaller tails both estimators show an increasing RMSE for the scale estimation due to an increasing variance. This results from the fact that in case of very small tails, single very large values can occur.

```{r MSE, echo=FALSE,warning=FALSE, fig.height=7,fig.width=8,fig.align="center", fig.cap="\\label{fig:MSE} RMSE for the estimation of tail (left column) and scale (right column) parameters via log-Moment estimator and MLE of a Mittag-Leffler sample with varying sample size n=25, 100, 200, varying tails on the x-axis and fixed scale equal to one, based on 5000 simulation runs.", out.width = '90%'}
load(here::here("article_springer", "article_springer_files", "ML_estim_simu","tbl_ML_estim.RData"))
  n.labs <-c("n=25", "n=100", "n=200")
  names( n.labs ) <-  c("25", "100", "200")
tbl_ML_estim %>% filter(method=="logMom" | method=="ML")%>%
  filter(n == "25" || n == "100" || n=="200") %>%
  filter(true.scale == "1") %>% 
  ggplot(aes(true.tail, rmse, colour = method, group=method)) + 
  geom_line() + geom_point() + 
  facet_wrap(c("n","param"), scales = "free", ncol = 2, labeller = labeller(n = n.labs)) + 
  scale_color_manual(values=c("red", "blue"),name = "Method", labels = c("log-Moment", "MLE")) + ggtitle("RMSE for log-Moment and MLE") +
   theme(plot.title = element_text(hjust = 0.5))

Since the Mittag-Leffler distribution is heavy-tailed, many researchers would intuitively give the highest importance to the tail behaviour of the distribution. Of course, one can also use established tail exponent estimators for the estimation of the parameter $\beta$ such as the Hill estimator [@hill1975simple] based on the $r+1$ upper order statistics
\begin{align}\label{eq:Hill}
H_{r,n}=\left[ \frac{1}{r} \sum_{i=0}^{r} \log \frac{X_{(i)}}{X_{(r+1)}}\right]^{-1}, 
\end{align}
where $X_{(1)} \geq X_{(2)} \geq \ldots \geq X_{(n)}$ denote the order statistics in decreasing order of a sequence $X_1,\ldots,X_n$. The Hill estimator only performs reliably well for distributions close to Pareto [see @Resnick97]. It heavily depends on the choice of $r$ in \eqref{eq:Hill}, and hence on the number of order statistics entering \eqref{eq:Hill}.  Figure \ref{fig:Hillplots} shows 100 Hill plots (grey thin lines) for simulated Mittag-Leffler data with tail parameter $\beta=0.8$ and scale $\sigma=1$, for a sample size of $n=200$ (left) and $n=1000$ (right). The dark lines are the means and the red dotted lines show the true tail
parameter. To deduce the correct tail parameter estimates from one of those Hill plots is virtually impossible. To estimate the tail parameter of a Mittag-Leffler distribution with the Hill estimator, the sample size has to be much larger and one has to choose a small tuning parameter $r$. In case that only events above a high threshold are recorded, $n=10000$ events are extremely unrealistic. There is a vast literature on bias-reduced versions of the Hill estimator and on the optimal choice of $r$, mainly based on imposed second-order regular variation conditions [see e.g. @beirlantBook, Ch.4.4 and 4.7 and references in it]. Also, there are hundreds other tail-index estimators not considered in this article, since this would go beyond the scope of our article. For example the moment estimator [@dekkers1989moment], the Pickands estimator [@pickands1975statistical], the QQ-estimator [@Kratz96], to name a few. A nice overview can be found in [@fedotenkov2018review]. However, we know from the Theorem in Section \ref{sec:scaling} that the inter-exceedance times are asymptotically Mittag-Leffler distributed, and hence it is unsurprising that general tail (and scale) estimation methods that do not take into account this structure are statistically less efficient [@kozubowski2001]. Finally, the Hill estimator completely fails in the important special case where the inter-arrival times are exponentially distributed and not heavy-tailed. For these reasons, we propose to use the log-Moment estimator or the MLE for Mittag-Leffler distributions.

@@

mean_n200<-hill_plot_values %>% group_by(x) %>% summarize(mean=mean(y))
p1<-hill_plot_values %>% ggplot() + geom_line(aes(x=x,y=y,group=rep),colour="lightgrey") +
  coord_cartesian(ylim = c(0.5, 1.5)) +
  geom_line(aes(x=x, y=mean),colour="black",data=mean_n200) + 
  ylab("Estimated tail") +
  geom_hline(yintercept=0.8, linetype="dashed", color = "red")+
  xlab("Order Statistics")+
  ylab("Hill-Estimator of Tail Index")

hill_plot_values<-data.frame(x=numeric(),y=numeric(),rep=numeric())
for  (m in 1:100){
  data_ml_2<-rml(n=10000,tail=0.8,scale=1)
  df<-hill_estim_range(k_min=30,k_max=200,data=data_ml_2,true_tail=0.8)
  df<-df %>% mutate(rep=m)
  hill_plot_values<-bind_rows(df,hill_plot_values)
}
mean_n1000<-hill_plot_values %>% group_by(x) %>% summarize(mean=mean(y))

p2<-hill_plot_values %>% ggplot() + geom_line(aes(x=x,y=y,group=rep),colour="lightgrey") +
  coord_cartesian(ylim = c(0.5, 1.5)) +
  geom_line(aes(x=x, y=mean),colour="black",data=mean_n1000) + 
  ylab("Estimated tail") +
  geom_hline(yintercept=0.8, linetype="dashed", color = "red")+
  xlab("Order Statistics")+
  ylab("Hill-Estimator of Tail Index")

@@

@@

```{r LRT_power, eval=FALSE, echo=FALSE,warning=FALSE, fig.align="center", fig.cap="\\label{Fig:LRT} Power for bootstrapped LRT for different sample sizes, varying tails and scale parameter equal to 1.", out.width='90%'}
load(here::here("article_springer","article_springer_files","ML_LRT_simu","estimates_df_LRT_boot1_01.RData"))
df_LRT_boot_01 %>% filter(n=="20" || n=="50" || n=="100" || n=="200") %>% ggplot(aes(x=true.tail,y=power)) +  geom_line(aes(y=power,group=factor(n),color=factor(n),linetype=factor(n)),size=0.5)+
  scale_linetype_manual(values = c("solid","dashed","twodash","dotted"),name  ="Sample size")+
   labs(y="Power",x= "True tail parameter")+
  scale_color_manual(values=c("red", "blue","darkgreen","black"), name  ="Sample size") +
  ggtitle("Power of Likelihood-Ratio Test")+
   theme(plot.title = element_text(hjust = 0.5))
```

# Inference on Exceedance times

@@

1. Extract the $K$ largest order statistics 
(i.e. the $K$ largest values, where e.g. $K = 300$) 
$X_{(1)}, \ldots, X_{(K)}$ together with their timestamps $T_{(1)}, \ldots, T_{(K-1)}$.

@@

To clarify Step 4: Recall that by the theorem, $\sigma_k / b(1/p_{\ell})$ is expected to stabilize 
around a constant as $k$ gets smaller. Since $b$ is regularly varying with parameter
$1/\beta$, we have $b(1/p_{\ell}) = p_{\ell}^{-1/\beta} / L(1/p_{\ell})$ for some slowly varying function $L$. 
Approximating $p_{\ell}$ by $\hat p_{\ell} := k / n$, we have 
$$\text{const} \approx \sigma_k / b(1/p_{\ell}) = \sigma_k \times p_{\ell}^{1/\beta} L(1/p_{\ell}) \approx \sigma_k \times k^{1/\beta} n^{-1/\beta} L(n/k)$$
Assuming that the variation of $L(n/k)$ is minor, we can hence see that 
$\sigma_k k^{1/\beta}$ stabilizes.

@@

: We approximated $p_{\ell}$, the probability that an event is larger than $l$, by its relative frequency. One could also approximate this tail probability by the GPD distribution fitted to the exceedances.

@@

We plot the tail parameter estimates $\hat \beta(k)$ against $k$ 
for each of the $m=100$ simulation runs. 
Thin grey lines represent individual simulation runs, 
and the thicker black line is their mean. 
Recall that $k$ is the index of the order statistics of $J_k$ at which the 
threshold $\ell$ is placed. 
<!-- Note that the Hill estimator is based on the waiting times, not on the  -->
<!-- inter-exceedance times, meaning that the estimation is based on more data.  -->
<!-- If one were to base the Hill-estimation only on the inter-exceedance times,  -->
<!-- the sample size would be too small and it would be impossible to deduce an  -->
<!-- estimate for the tail parameter from the plots (compare Figure \ref{fig:Hillplots}).  -->

>>>>

  A new model for the _inter-exceedance times_ of events 
  above high thresholds is proposed. For high thresholds and 
  infinite-mean waiting times, it is shown that the times between threshold 
  crossings are Mittag-Leffler distributed, and thus form a "fractional 
  Poisson Process" which generalizes the standard Poisson Process of threshold
  exceedances.
  Graphical means of estimating model parameters and assessing 
  model fit are provided. 
  The inference method is
  applied to an empirical bursty time series, and it is shown how the memory of the
  Mittag-Leffler distribution affects prediction of the time until
  the next extreme event."

@@

We review the marked renewal process in Section 2.1, and 
derive a scaling limit theorem for inter-exceedance times in Section 2.2.
We give a statistical procedure to estimate model parameters via stability plots in Section 3.1 and 3.4, but first we need to discuss inference for the Mittag-Leffler distribution in Section 3.2 as well as a Likelihood-ratio test to guide the choice, whether a Mittag-Leffler or an exponential distribution fits better to the inter-exceedance times. A simulation study of the effectiveness of our statistical procedure is given in Section 4. In Section 5 we apply our method to a real data set.  In Section 6, we discuss the memory property of the  Mittag-Leffler distribution, and how it affects the predictive distribution for the time until the next threshold crossing event. Finally we close with a discussion and conclusion in Section 7.
For all statistical computations we have used `R` [@R] and the package `CTRE`
[@CTRE]. Source code for simulations and figures generated in this 
manuscript is available online at <https://github.com/strakaps/bursty-POT>.

# Probabilistic Model

## Continuous Time Random Exceedances (CTRE) {#sec:CTRE}

@@

: Let $(W,J), (W_1, J_1), (W_2, J_2), \ldots$ be iid pairs of random variables, where the $W_k > 0$ are interpreted as the *waiting times* 
and $J_k \in \mathbb{R}$ as the *event magnitudes*. If $W$ and $J$ are independent, the Marked Renewal Process is said to be *uncoupled*.

@@

: Given a threshold $\ell \in (x_L, x_R)$, 
consider the stopping time 
$$\tau(\ell) := \min\{k: J_k > \ell\},\quad \ell \in (x_L, x_R).$$
Define the pair of random variables $(X(\ell), T(\ell))$ via 
$$X(\ell) = J_{\tau(\ell)} - \ell, \quad 
T(\ell) = \sum_{k=1}^{\tau(\ell)} W_k.$$
By restarting the MRP at $\tau(\ell)$, inductively define the
two iid sequences $T(\ell,n)$ and $X(\ell, n)$, $n \in \mathbb N$, 
called the "inter-exceedance times" (IETs) and the "exceedances", respectively.
The pair sequence $(X(\ell, n), T(\ell, n))_{n \in \mathbb N}$ is called 
a Continuous Time Random Exceedance model (CTRE). 
If the underlying MRP is uncoupled, then the CTRE is also called 
uncoupled.

@@

## Scaling limit of Exceedance Times {#sec:scaling}

@@

$$
\lim_{x \to \infty} \frac{f(\lambda x)}{f(x)} = \lambda^\rho \quad \text{ for all } \lambda > 0.
$$

For more details on regular variation and stable limit theorems, we recommend the book by @MeerschaertSikorskii.

@@

: For the magnitudes $J_k$, let assumption \eqref{assumptionJs} hold. Furthermore, let the waiting times $W_k$ be in the domain of attraction of a positively skewed sum-stable law with stability parameter $0 < \beta < 1$; more precisely,
\begin{align} \label{eq:stability}
\frac{W_1 + \ldots + W_n}{b(n)} \Rightarrow D, 
\quad n \to \infty
\end{align}
for a function $b(n)$ which is regularly varying at $\infty$ with parameter $1/\beta$, and where $\mathbf E[\exp(-sD)] = \exp(-s^\beta)$, $s > 0$. Write $p_{\ell} := \mathbf P(J > \ell)$. Then the weak convergence
$$
\frac{T(\ell)} {b(1/p_{\ell})} \Rightarrow  Z_\beta \quad \text{ as } \quad \ell \uparrow x_R
$$
holds, where the Mittag-Leffler random variable $Z_\beta$ is defined on 
the positive real numbers via 
$$
\mathbf E[\exp(-sZ_\beta)] = \frac{1}{1+s^\beta}.
$$

*Proof of Theorem:* 	Due to assumption \eqref{assumptionJs}
\begin{align*}
& & P\left(\frac{J_1 \vee \ldots \vee J_{\lfloor c \rfloor}-d(c)}{a(c)} \leq x\right) & \rightarrow F(x;\xi) \\
& \Longleftrightarrow & F_J(xa(c)+d(c))^{\lfloor c \rfloor} & \rightarrow F(x;\xi) \\
& \Longleftrightarrow &  \lfloor c \rfloor \log F_J(\ell_c) & \rightarrow \log F(x;\xi)
\end{align*}
as $c \rightarrow \infty$, for any $x$ from 
the support of $F(x;\xi)$,	with $\ell_c:=xa(c)+d(c)$. Furthermore, since $\log(1-x) \sim -x$ for small $x$ it follows
\begin{align*}
c \cdot p(\ell_c) \rightarrow -\log F(x;\xi) \quad  \text{ as } c \rightarrow \infty,
\end{align*}
with $p(\ell_c):= 1-F(\ell_c)$. Due to $\tau(\ell_c) \sim {\rm Geo}(p(\ell_c))$ and above equation,
	it follows that $\tau(\ell_c)/c$ converges to an exponential random variable,
\begin{align*}
	\frac{\tau(\ell_c)}{c} \Rightarrow E \quad  \text{ as } c \rightarrow \infty
\end{align*}
	 with inverse mean $\lambda:=-\log F(x;\xi)$. Due to
\begin{align*}
	S(c):= \sum_{i=1}^{\lfloor c \rfloor} \frac{W_i}{b(c)} \Rightarrow D \quad \text{ as } c \rightarrow \infty,
\end{align*}
    it follows with Gnedenko's transfer theorem (see @gnedenko1983limit), that
\begin{align*}
	\sum_{i=1}^{ \tau(\ell_c)} \frac{W_i}{b(\ell_c)}  \Rightarrow Z \quad \text{ as }  c \rightarrow \infty, 
\end{align*}
where the distribution of $Z$ has the characteristic function
\begin{align*}
\Psi(s) &= \int_0^{\infty} (\Phi_D(s))^{y} F_E(dy) = \frac{1}{1-\log(\Phi_D(s))/\lambda}\\
&= \frac{1}{1+s^{\beta}\lambda^{-1}} = \frac{1}{1+(s \lambda^{-1/\beta})^{\beta}}
\end{align*}
    where $\Phi_D(s) = \exp(-s^\beta)$ is the Laplace transform of $D$ 
    and  $F_E(dy)$ is the distribution function of $E$. Hence, $Z$ is Mittag-Leffler distributed with scale parameter $\lambda^{-1/\beta}$. Rewriting
\begin{align*}
    \sum_{i=1}^{ \tau(\ell_c)} \frac{W_i}{b(1/p(\ell_c))} = \left(\sum_{i=1}^{ \tau(\ell_c)} \frac{W_i}{b(c) }\right) \frac{b(c)}{b(1/p(\ell_c))}  
\end{align*}
	and
\begin{align*}
	\frac{b(c)}{b(1/p(\ell_c))} = \frac{b(c)}{b(c/(p(\ell_c)c))} \sim (p(\ell_c) \cdot c)^{1/\beta} \rightarrow \lambda^{1/\beta} \text{ as } c \rightarrow \infty,
\end{align*}
	the second factor converges to $\lambda^{1/\beta}$, and it follows that
	 $Z \sim {\rm ML}(\beta,1)$. Since $c \rightarrow \infty$ is equivalent to $\ell \uparrow x_R$, the assertion follows with $\ell:=\ell_c$ and $p_\ell:=p(\ell_c)$.
\qed

@@

: If $\beta = 1$, the result of the Theorem above is standard, see  
Equation (2.2) in @Gut1999. In @Anderson1987  a similar result is shown with a different choice of scaling constant. [@MeerschaertStoev08] proved a limit theorem for the maxima of iid random variables separated by infinite mean waiting times. They have also shown that the hitting time of the limit process is Mittag-Leffler distributed. [@Basrak2014] described the asymptotic distribution, under similar assumptions, of all upper orders statistics using point processes.

@@

# Statistical Inference

## Mittag-Leffler Stability Plots

We assume a dataset of the form $\{(t_i, x_i), i \in I\}$,
where $t_i$ are timestamps and $x_i$ are magnitudes.
In this article, we focus on modelling the distribution of threshold
*exceedance times* rather than the *exceedances* themselves. For the latter, 
see the POT approach [@smith1984threshold;
@leadbetter1991basis; @davison1990models; @beirlantBook Ch.5.3;
@embrechts2013modelling Ch. 6.5.; @ColesBook Ch.4.].
With the previous section setting the stage, we assume that only the large
magnitudes (or "shocks") follow a MRP.  The smaller magnitudes are possibly
nuisance data that may be irrelevant with respect to the modelling of extremes
and their waiting times. 
Accordingly, we assume a threshold $\ell$, and discard all
data where $x_i \le \ell$, yielding the thresholded dataset
$\{(t^{(\ell)}_i, x^{(\ell)}_i): i \in I(\ell)\}$, with $x^{(\ell)}_i=x_i-\ell$, where $I(\ell):=\left\{i \in I: x_i \ge \ell \right\} \subseteq I$.

However, this raises the question of how high or low the threshold should be
chosen. Recall that with the POT method for the modelling of extremes, the
distribution of the thresholded exceedances converges to a GPD, and the
threshold must be chosen high enough so that the exceedances fit a GPD well. 
The threshold choice translates into a
bias-variance trade-off: low thresholds yield more data to fit a GPD (low 
variance), but the distribution of exceedances may deviate from a GPD 
distribution (high bias); and high thresholds present the opposite scenario.
It is understood, however, that there is a *range* of threshold values that
ought to yield similar GPD parameter estimates. In other words, plots of
threshold vs. parameter estimate ought to exhibit that parameter estimates
are *stable* with respect to the choice of threshold.

For inter-exceedance *times* (IETs), the same idea applies. 
By the Theorem, the inter-exceedance times converge to a Mittag-Leffler
distribution (MLD), and importantly, the tail parameter $\beta$ is 
*independent* of (or *stable* with respect to) the choice of threshold.
For high thresholds, the IET distribution may fit an MLD well, but few
IETs may be left for a fit (high variance). 
For low thresholds, their distribution may be not well represented
by an MLD; or, too many events may be interpreted as 
shocks, resulting in a biased estimate of the parameters of the limiting MLD.
If we had MRP data $\{(W_i, J_i): i \in I\}$ as defined in Section 
\ref{sec:CTRE}, we could apply any tail (and scale) estimator such as e.g. Hill's 
estimator [@hill1975simple] to the $W_i$ and thus infer the distribution of 
inter-exceedance times. However, in real-world datasets, the MRP assumption is too 
strong for low thresholds. The unknown data generating process is likely more 
complex, with dependencies or multiple data generating mechanisms at low thresholds 
and structures that may vanish for higher thresholds. Fitting only the IETs means 
that the iid assumption applies to exceedances and IETs only.

For high thresholds where few IETs are left, estimating a tail parameter
by fitting a MLD has advantages over e.g. the Hill estimator: 
since the underlying distribution of IETs *is* a MLD, 
the fits are more accurate. Figure \ref{fig:Hillplots} shows Hill plots for 200 IETs, for two Mittag-Leffler datasets of sizes 200 and 10000 (and an irrelevant distribution of magnitudes).
Clearly the variance is too high to give useful estimates. For corresponding fits based on the assumption of a Mittag-Leffler distribution, see Figure \ref{Fig:TailSimu}, top row, yielding more reliable predictions 
(more discussion further below).

@@

mean_n200<-hill_plot_values %>% group_by(x) %>% summarize(mean=mean(y))
p1<-hill_plot_values %>% ggplot() + geom_line(aes(x=x,y=y,group=rep), alpha=0.1) +
  coord_cartesian(ylim = c(0.5, 1.5)) +
  geom_line(aes(x=x, y=mean),colour="black",data=mean_n200) + 
  ylab("Estimated tail") +
  geom_hline(yintercept=0.8, linetype="dashed", color = "red")+
  xlab("Order Statistics")+
  ylab("Hill-Estimator of Tail Index")

hill_plot_values<-data.frame(x=numeric(),y=numeric(),rep=numeric())
for  (m in 1:100){
  data_ml_2<-rml(n=10000,tail=0.8,scale=1)
  df<-hill_estim_range(k_min=30,k_max=200,data=data_ml_2,true_tail=0.8)
  df<-df %>% mutate(rep=m)
  hill_plot_values<-bind_rows(df,hill_plot_values)
}
mean_n10000<-hill_plot_values %>% group_by(x) %>% summarize(mean=mean(y))

p2<-hill_plot_values %>% ggplot() + geom_line(aes(x=x,y=y,group=rep), alpha=0.1) +
  coord_cartesian(ylim = c(0.5, 1.5)) +
  geom_line(aes(x=x, y=mean),colour="black",data=mean_n10000) + 
  ylab("Estimated tail") +
  geom_hline(yintercept=0.8, linetype="dashed", color = "red")+
  xlab("Order Statistics")+
  ylab("Hill-Estimator of Tail Index")

@@

## Fitting Mittag-Leffler Distributions {#sec:ML}

The Maximum Likelihood Estimator (MLE) for the MLD is not straightforward 
to implement, since the MLD density admits no closed form analytical expression.
In the R Package `MittagLeffleR` [@MittagLeffleR], MLE is implemented via numerical optimization. The MLE slightly outperforms the log-Moment estimator regarding bias and RMSE for big enough sample sizes, but is computationally very intensive. Figure \ref{fig:MSE} shows that both estimators perform well, even for small sample sizes. For smaller tails both estimators show an increasing RMSE for the scale estimation due to an increasing variance. This results from the fact that in case of very small tails, single very large values can occur.

```{r MSE, echo=FALSE,warning=FALSE, fig.height=7,fig.width=8,fig.align="center", fig.cap="\\label{fig:MSE} RMSE for the estimation of tail (left column) and scale (right column) parameters via log-Moment estimator and MLE of a Mittag-Leffler sample with varying sample size n=25, 100, 200, varying tails on the x-axis and fixed scale equal to one, based on 5000 simulation runs.", out.width = '90%'}
load(here::here("article_springer", "article_springer_files", "ML_estim_simu","tbl_ML_estim.RData"))
  n.labs <-c("n=25", "n=100", "n=200")
  names( n.labs ) <-  c("25", "100", "200")
tbl_ML_estim %>% filter(method=="logMom" | method=="ML")%>%
  filter(n == "25" | n == "100" | n=="200") %>%
  filter(true.scale == "1") %>% 
  ggplot(aes(true.tail, rmse, colour = method, group=method)) + 
  geom_line() + geom_point() + 
  facet_wrap(c("n","param"), scales = "free", ncol = 2, labeller = labeller(n = n.labs)) + 
  scale_color_manual(values=c("red", "blue"),name = "Method", labels = c("log-Moment", "MLE")) + ggtitle("RMSE for log-Moment and MLE") +
   theme(plot.title = element_text(hjust = 0.5))

With the above comparisons of estimators for the MLD in mind, 
we propose to use the log-Moment estimator, or the MLE when computational 
resources are not an issue.

## Weighing the evidence for non-exponential inter-arrival times

@@

```{r LRT_power, echo=FALSE,warning=FALSE, fig.align="center", fig.cap="\\label{Fig:LRT} Power for bootstrapped LRT for different sample sizes, varying tails and scale parameter equal to 1.", out.width='90%'}
load(here::here("article_springer","article_springer_files","ML_LRT_simu","estimates_df_LRT_boot1_01.RData"))
df_LRT_boot_01 %>%  ungroup() %>% filter(n=="20" | n=="50" | n=="100" | n=="200") %>% ggplot(aes(x=true.tail,y=power)) +  geom_line(aes(y=power,group=factor(n),color=factor(n),linetype=factor(n)),size=0.5)+
  scale_linetype_manual(values = c("solid","dashed","twodash","dotted"),name  ="Sample size")+
   labs(y="Power",x= "True tail parameter")+
  scale_color_manual(values=c("red", "blue","darkgreen","black"), name  ="Sample size") +
  ggtitle("Power of Likelihood-Ratio Test")+
   theme(plot.title = element_text(hjust = 0.5))
```

## Algorithm for the inference on inter-exceedance times

@@

1. Extract the $K$ largest order statistics 
(i.e. the $K$ largest values, where e.g. $K = 300$) 
$x_{(1)}, \ldots, x_{(K)}$ together with their timestamps $t_{(1)}, \ldots, t_{(K-1)}$.

@@

To clarify Step 4: Recall that by the theorem, $\sigma_k / b(1/p_{\ell})$ is expected to stabilize 
around a constant as $k$ decreases. Since $b$ is regularly varying with parameter
$1/\beta$, we have $b(1/p_{\ell}) = p_{\ell}^{-1/\beta} / L(1/p_{\ell})$ for some slowly varying function $L$. 
Approximating $p_{\ell}$ by $\hat p_{\ell} := k / n$, we have 
$$\text{const} \approx \sigma_k / b(1/p_{\ell}) = \sigma_k \times p_{\ell}^{1/\beta} L(1/p_{\ell}) \approx \sigma_k \times k^{1/\beta} n^{-1/\beta} L(n/k)$$
Assuming that the variation of $L(n/k)$ is minor, we can hence see that 
$\sigma_k k^{1/\beta}$ stabilizes.

@@

: We approximated $p_{\ell}$, the probability that an event is larger than $l$, by its relative frequency. One could also approximate this tail probability via the GPD distribution fitted to the exceedances.

@@

We plot the tail parameter estimates $\hat \beta(k)$ against $k$ 
for each of the $m=100$ simulation runs. 
Thin grey lines represent individual simulation runs, 
and the thicker black line is their mean. 
Recall that $k$ is the index of the order statistics of $J_k$ at which the 
threshold $\ell$ is placed.
